---
output:
  md_document:
    variant: markdown_github
---

# curtail: an R package for test curtailment

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/README-figures/README-",
  dpi = 124
)
```

A curtailed test is a variable-length test, which allows for early stopping of item administration when further items are unlikely or unable to change the final (classification) decision. The package allows for creating and assessing deterministically and stochastically (based on empirical proportions) curtailed tests. In future versions of the package, model-based stochastic curtailment may be added. The package allows for test curtailment through the creation of look-up tables, providing item-specific cut-off values for early stopping of item  administration. In addition, the package allows for assessing efficiency (number of items administered to arrive at a final decision) and accuracy (concordance between the final decision based on curtailed and full-length test).


## Example

To get a first impression of how curtail works, we take a dataset of item scores on a 20-item test:

```{r, results = FALSE}
library(curtail)
head(itemscores)
```

First, we apply deterministic curtailment on all observations, using a cut-off value of 17:

```{r}
tmp <- Curtail(itemscores, Xstar = 17)
summary(tmp$current.item)
```

Perhaps we can further reduce test length through stochastic curtailment. We use the first 500 observations for training and the next 500 observations for testing:

```{r}
tmp <- stochCurtail(itemscores[1:500,], dataset.test = itemscores[501:1000,], 
                    Xstar = 17)
summary(tmp$current.item)
```

Indeed, we were able to reduce the number of items administered somewhat. However, this also yielded 4 incorrect decisions (missed at-risk observations). Perhaps we want to inspect test length distributions for at-risk or not-at-risk observations only:

```{r}
hist(tmp$current.item[tmp$SCrisk], xlab = "Number of items administered", 
     main = "Test lengths for at-risk observations")
hist(tmp$current.item[tmp$SCnorisk], xlab = "Number of items administered", 
     main = "Test lengths for not-at-risk observations")
```

If we want to obtain tables with item-specific cutoff values, we can use the Table and stochTable functions:

```{r}
Table(itemscores, Xstar = 17)
stochTable(itemscores, Xstar = 17)
```

Perhaps we also want to assess the accuracy of decisions based on stochastic curtailment, using leave-on-out cross validation. This can be done using the stochCurtailXval function:

```{r}
stochCurtailXval(itemscores[1:250,], Xstar = 17)
```

